---
title: "Catch Original Quarto document"
format: html
editor: visual
---

```{r}
library(dplyr)

library(tidyr)

library(readr)

#| message: false


```

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Data sources

Read in the data

Explore the data

```{r}
## Prints the column names of my data frame
colnames(catch_original)

## First 6 lines of the data frame
head(catch_original)

## Summary of each column of data
summary(catch_original)

## Prints unique values in a column (in this case, the region)
unique(catch_original$Region)

## Opens data frame in its own tab to see each row and column of the data (do in console)
View(catch_original)
```

## Clean and reshaping data

Using select() function

```{r}
catch_data <- catch_original %>%
    select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)
```

```{r}
catch_data <- catch_original %>%
    select(-All,-notesRegCode)
```

Check column types

```{r}
dplyr::glimpse(catch_data)
```

Replace values in a column with mutate()

```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))
```

```{r}
head(catch_clean)
```

```{r}
i <- which(is.na(catch_clean$Chinook))
i
```

```{r}
catch_data[i,]
```

```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = if_else(condition = Chinook == "I", 
                             true = "1", 
                             false = Chinook),
           Chinook = as.numeric(Chinook))

##check
catch_clean[i, ]
```

Reshape data with pivot_longer() and pivot_wider()

```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch"
    )

head(catch_long)
```

```{r}
catch_wide <- catch_long %>%
    pivot_wider(names_from = species,
                values_from = catch)

head(catch_wide)
```

Rename columns rename()

```{r}
catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```

Replace values in a column with mutate()

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
    select(-catch_thousands)

head(catch_long)
```

Summary stats using group_by() and summarize()

```{r}
unique(catch_original$Region)
```

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)
```
```{r}
## for example:
catch_year_sp <- catch_long %>%
    group_by(Year, species) %>%
    summarize(total_year = sum(catch, na.rm = T))
```

Filtering rows using filter()

Sort data using arrange()

Split and combine values in columns with separate() and unite()
